---
title: "Mapping BC health geographies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping BC health geographies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8
)
library(ggplot2)
library(islandbrand)
# Resolve the brand font first, as the other articles do. Unresolved, the
# theme names BC Sans, and where it is not installed the PDF device that
# R CMD check runs this code on stops with "invalid font type".
islh_setup(format = "plots", quiet = TRUE)
```

The two packages divide mapping work by responsibility. `islandepi` retrieves
and standardizes population and health-geography data. `islandbrand` controls
the visual presentation of the finished map.

## See the map functions work

This first example uses four simple polygons created in the vignette. It does
not contact the BC Data Catalogue, so the website can display a real map while
remaining reproducible.

```{r map-example-data}
demo_map <- data.frame(
  geography_name = c("North", "Central", "West", "South"),
  population = c(12000, 34000, 18000, 52000),
  wkt = c(
    "POLYGON ((0 1, 1 1, 1 2, 0 2, 0 1))",
    "POLYGON ((1 1, 2 1, 2 2, 1 2, 1 1))",
    "POLYGON ((0 0, 1 0, 1 1, 0 1, 0 0))",
    "POLYGON ((1 0, 2 0, 2 1, 1 1, 1 0))"
  )
)
demo_map <- sf::st_as_sf(demo_map, wkt = "wkt", crs = 3005)
demo_map[c("geography_name", "population")]
```

```{r map-example, fig.alt="Four-area demonstration choropleth, with the southern area darkest because it has the largest illustrative population."}
ggplot(demo_map, aes(fill = population)) +
  geom_sf(colour = islh_hex("grey", 80), linewidth = 0.4) +
  scale_fill_islh_b(
    n.breaks = 4
  ) +
  coord_islh_map() +
  labs(
    title = "Population by area",
    subtitle = "Demonstration geometry and values",
    fill = "Population",
    caption = islh_caption(
      source = "Simulated",
      extracted = "2026-09-04"
    )
  ) +
  theme_islh_map()
```

The same three presentation functions work with downloaded LHA, HSDA, CHSA or
health-authority boundaries. The next section replaces `demo_map` with real
catalogue data.

## Retrieve and join the data

The following example requests total population and LHA boundaries for Island
Health. It is shown but not run here, because the catalogue can be unavailable
even when both packages are working correctly.

```{r retrieve, eval=FALSE}
library(dplyr)
library(ggplot2)
library(islandepi)
library(islandbrand)

population_extract <- islh_bc_population(
  geography = "lha",
  years = 2025,
  sex = "T"
)

population <- population_extract |>
  summarise(
    population = sum(population),
    .by = c(geography, geography_code, year, estimate_type)
  )

boundaries <- islh_bc_geography("lha")

map_data <- left_join(
  boundaries,
  population,
  by = join_by(geography, geography_code),
  relationship = "one-to-one",
  na_matches = "never"
)

if (anyNA(map_data$population)) {
  stop("One or more Island Health boundaries did not match population data.")
}
```

The population resource covers all of BC, while the default boundary object
contains only Island Health. Extra provincial population rows are expected to
be discarded. The one-to-one relationship rejects duplicate codes, and the row
check catches any Island Health boundary without a population match.

`islandbrand` ships the result of that join for 2025, simplified for display, as
`islh_example_lha()`. The rest of this article draws from it, so every map
below is real:

```{r example-lha}
map_data <- islh_example_lha()
map_data
```

## Clip the boundaries to land

The catalogue's LHA boundaries already follow the coastline. Other layers, such
as census or custom boundaries, often extend over water, and drawn unclipped
Island Health becomes a blob across the Salish Sea. Clip those to a land layer
first, then simplify. This needs a land layer of your own, so it is not run
here:

```{r clip, eval=FALSE}
map_data <- map_data |>
  sf::st_intersection(bc_land) |>
  rmapshaper::ms_simplify(keep = 0.05, keep_shapes = TRUE)
```

Use `rmapshaper::ms_simplify()` rather than `sf::st_simplify()`. It preserves
shared borders, so adjacent areas do not end up with slivers between them, and
`keep_shapes = TRUE` keeps the small islands. Neither package is a dependency
of `islandbrand`; install them in the analysis project.

## Draw the choropleth

Three functions carry the presentation:

* `theme_islh_map()` starts from `ggplot2::theme_void()` and adds the brand
  text hierarchy, so nothing but the data draws.
* `coord_islh_map()` locks the map to BC Albers (EPSG:3005) and removes the
  graticule.
* `scale_fill_islh_b()` places a continuous value into branded colour steps.

```{r choropleth, fig.alt="Choropleth of Island Health's 14 local health areas by 2025 population. Greater Victoria is in the darkest band; Vancouver Island West and Vancouver Island North are in the lightest."}
base_map <- ggplot(map_data, aes(fill = population)) +
  geom_sf(colour = islh_hex("grey", 80), linewidth = 0.25) +
  scale_fill_islh_b(breaks = c(25000, 50000, 100000, 200000)) +
  coord_islh_map() +
  labs(
    title = "Population by local health area",
    subtitle = "Island Health, 2025 estimates",
    fill = "Population",
    caption = islh_caption(
      source = "BC Stats and BC Data Catalogue",
      extracted = "2026-09-24",
      boundary = "Local health area boundaries, simplified for display",
      width = 60
    )
  ) +
  theme_islh_map()

base_map
```

Population is skewed: Greater Victoria has more than a hundred times the
people of Vancouver Island West. Breaks that double (25,000, 50,000, 100,000,
200,000) keep the small areas from falling into one band.

Always set the projection. Latitude and longitude (EPSG:4326) stretches
Vancouver Island sideways and misrepresents how large each area is. BC Albers
is what the provincial layers already use.

The legend abbreviates large counts, so `25000` prints as `25K`. A bottom
legend has little room between breaks, and full comma-separated numbers
overlap once a map has more than three or four bins. Pass
`labels = scales::label_comma()` when a figure needs exact numbers, along with
`key_width = 16` to lengthen the bar.

For a categorical map, use `scale_fill_islh()` instead. A map should retain a
text label or table alternative when colour is the only visual distinction.

## Put the legend in the empty corner

Island Health runs northwest to southeast, which leaves an empty corner at the
top right of a full-Island map, over the mainland the map does not draw.
`coord_sf()` fixes the aspect ratio, so a wider figure cannot stretch the map
to fill the space a bottom legend takes. Moving the legend inside the panel
gives the map roughly a fifth more room:

```{r legend-inside, fig.alt="The same population map with its legend placed in the empty top-right corner, above the east coast of Vancouver Island."}
base_map + theme_islh_map(legend = "inside")
```

Adjust `legend_inside` when the map shows only part of the Island. It takes
fractions of the panel, measured from the bottom left, for the legend's
bottom-left corner.

## Map rates, and classify on purpose

Map age-standardized rates rather than counts, and name the standard
population in the caption. A count map of local health areas mostly shows
where people live.

Choose the bin breaks deliberately. `classInt::classIntervals()` offers
quantile, Jenks and equal-interval methods; quantile suits comparisons between
areas, equal interval suits the same indicator across time. Then hard-code the
breaks when a report maps the same indicator more than once.

The cases here are simulated at about 90 per 100,000, so the map shows the
method, not a real pattern:

```{r rate-map, fig.alt="Choropleth of simulated crude rates per 100,000 by local health area, in four fixed bands."}
set.seed(2026)
map_data$cases <- stats::rpois(nrow(map_data), map_data$population * 90 / 1e5)
map_data$rate <- map_data$cases / map_data$population * 1e5

ggplot(map_data, aes(fill = rate)) +
  geom_sf(colour = islh_hex("grey", 80), linewidth = 0.25) +
  scale_fill_islh_b(breaks = c(80, 90, 100)) +
  coord_islh_map() +
  labs(
    title = "Crude rate by local health area",
    subtitle = "Per 100,000; simulated cases",
    fill = "Rate per 100,000"
  ) +
  theme_islh_map(legend = "inside")
```

Bins recalculated from each year's data make an unchanged pattern look like a
change.

This example maps a crude rate because the example data holds only each
area's total population, with no age groups. Crude rates of areas with
different age structures are not comparable: an older area has more cases of
most conditions without any difference in risk. For a published map, count
cases and population by age group and standardize, with
`islandepi::islh_dsr_joined()`, one area at a time. The map code is the same;
only the column, the title and the caption change:

```{r standardized-rates, eval=FALSE}
standard <- islandepi::islh_standard_population(
  "canada_2011",
  age_breaks = c(0, 20, 45, 65, 85)
)

# cases_by_age and population_by_age: one row per area and age_group.
standardized <- do.call(rbind, lapply(map_data$geography_code, function(code) {
  rate <- islandepi::islh_dsr_joined(
    cases_by_age[cases_by_age$geography_code == code, ],
    population_by_age[population_by_age$geography_code == code, ],
    standard,
    by = "age_group",
    standard_id = "Canada 2011, five age groups"
  )
  data.frame(geography_code = code, rate)
}))
```

Name the standard population and its age groups in the caption, such as
"Age-standardized to the 2011 Canadian population".

Five to seven bins is the ceiling. Beyond that nobody can match a swatch back
to the map.

## Keep suppressed, zero and missing apart

Suppressed, zero and no data are three different statements and must read as
three different things. In this simulation, Vancouver Island West has 1 case,
so its count is suppressed. Suppose also that Cowichan Valley West had no cases
and that Vancouver Island North's data had not arrived:

```{r status-data}
status_data <- map_data
status_data$cases[status_data$geography_name == "Cowichan Valley West"] <- 0
status_data$cases[status_data$geography_name == "Vancouver Island North"] <- NA

status_data$status <- ifelse(
  is.na(status_data$cases),
  "No data",
  ifelse(
    status_data$cases == 0,
    "No cases",
    ifelse(status_data$cases < 5, "Suppressed: 1 to 4 cases", "Rate shown")
  )
)
status_data$rate <- ifelse(
  status_data$status %in% c("Rate shown", "No cases"),
  status_data$cases / status_data$population * 1e5,
  NA
)
table(status_data$status)
```

Draw the rates in their bands and the other three as a separate layer with its
own key. Each takes a lightness of its own, so a greyscale print still tells
them apart, and no data also has a dashed outline:

```{r status-map, eval=requireNamespace("ggnewscale", quietly = TRUE), fig.height=5.4, fig.alt="The simulated rate map. Eleven areas are in rate bands. Cowichan Valley West is pale blue, labelled No cases. Vancouver Island West is mid grey, labelled Suppressed: 1 to 4 cases. Vancouver Island North is white with a dashed outline, labelled No data."}
shown <- status_data[status_data$status == "Rate shown", ]
other <- status_data[status_data$status != "Rate shown", ]
statuses <- c("No cases", "Suppressed: 1 to 4 cases", "No data")
other$status <- factor(other$status, levels = statuses)

ggplot() +
  geom_sf(
    data = shown,
    aes(fill = rate),
    colour = islh_hex("grey", 80),
    linewidth = 0.25
  ) +
  scale_fill_islh_b(name = "Rate per 100,000", breaks = c(80, 90, 100)) +
  ggnewscale::new_scale_fill() +
  geom_sf(
    data = other,
    aes(fill = status, linetype = status),
    colour = islh_hex("grey", 40),
    linewidth = 0.35
  ) +
  scale_fill_manual(
    name = NULL,
    values = stats::setNames(
      c(islh_hex("blue", 90), islh_hex("grey", 80), "white"),
      statuses
    ),
    drop = FALSE
  ) +
  scale_linetype_manual(
    name = NULL,
    values = stats::setNames(c("solid", "solid", "22"), statuses),
    drop = FALSE
  ) +
  coord_islh_map() +
  labs(
    title = "Crude rate by local health area",
    subtitle = "Per 100,000; simulated; counts of 1 to 4 suppressed"
  ) +
  theme_islh_map() +
  # Two keys side by side are wider than the map; stack them.
  theme(legend.box = "vertical", legend.box.just = "left")
```

A zero is a result: here it is the palest blue, below the lowest band, rather
than grey. For a hatch rather than a flat fill, `ggpattern::geom_sf_pattern()`
does the same job as the second layer.

## Show the estimates and their intervals

A classified map cannot show how certain each rate is. Southern Gulf Islands
and Cowichan Valley South share the top band, but the first rests on 23 cases
and could be anywhere from about 76 to 180 per 100,000, while the second
rests on 67. When precision affects how the map should be read, put a table
beside it with each estimate and its interval. The table is also
the accessible alternative to the map: it says in words what the colours say.

This uses the exact Poisson interval, written out in base R.
`islandepi::islh_crude_rate(cases, population, method = "exact")` returns
the same values:

```{r rate-table, eval=requireNamespace("gt", quietly = TRUE)}
exact_interval <- function(cases, population, per = 1e5, level = 0.95) {
  alpha <- 1 - level
  lower <- ifelse(cases == 0, 0, stats::qchisq(alpha / 2, 2 * cases) / 2)
  upper <- stats::qchisq(1 - alpha / 2, 2 * cases + 2) / 2
  list(lower = lower / population * per, upper = upper / population * per)
}

areas_table <- sf::st_drop_geometry(status_data)
interval <- exact_interval(areas_table$cases, areas_table$population)
one_decimal <- function(x) formatC(x, format = "f", digits = 1)

areas_table$estimate <- ifelse(
  areas_table$status %in% c("Rate shown", "No cases"),
  paste0(
    one_decimal(areas_table$rate),
    " (",
    one_decimal(interval$lower),
    " to ",
    one_decimal(interval$upper),
    ")"
  ),
  areas_table$status
)
areas_table$cases_shown <- ifelse(
  areas_table$status %in% c("Rate shown", "No cases"),
  format(areas_table$cases, big.mark = ","),
  ifelse(areas_table$status == "No data", "No data", "1 to 4")
)

rate_table <- data.frame(
  "Local health area" = areas_table$geography_name,
  Cases = areas_table$cases_shown,
  Population = format(areas_table$population, big.mark = ","),
  "Rate per 100,000 (95% CI)" = areas_table$estimate,
  check.names = FALSE
)

islh_gt(
  rate_table,
  title = "Crude rate by local health area",
  source_note = paste(
    "Simulated cases. 95% confidence intervals, exact Poisson method.",
    "Counts of 1 to 4 suppressed."
  )
)
```

Vancouver Island North's population is known, but without its cases there is
no rate and no interval. The interval for Cowichan Valley West shows that no
cases in 7,329 people is consistent with a rate of up to about 50 per 100,000.

## Insets for the urban areas

Vancouver Island North and West dominate the page while Greater Victoria, where
most of the population lives, shrinks to a few small polygons. Build the dense
areas as their own panels and combine them:

```{r inset, eval=requireNamespace("patchwork", quietly = TRUE), fig.width=8, fig.alt="The population map with a dark rectangle around Greater Victoria and the Saanich Peninsula, beside an enlarged panel of that rectangle."}
inset_x <- c(1170000, 1215000)
inset_y <- c(375000, 420000)

victoria <- base_map +
  coord_islh_map(xlim = inset_x, ylim = inset_y) +
  labs(title = NULL, subtitle = NULL, caption = NULL) +
  theme_islh_map(legend = "none") +
  theme(
    panel.border = element_rect(
      colour = islh_hex("grey", 20),
      fill = NA,
      linewidth = 0.8
    )
  )

# The rectangle on the overview shows where the enlarged panel is.
overview <- base_map +
  annotate(
    "rect",
    xmin = inset_x[1],
    xmax = inset_x[2],
    ymin = inset_y[1],
    ymax = inset_y[2],
    fill = NA,
    colour = islh_hex("grey", 20),
    linewidth = 0.5
  )

patchwork::wrap_plots(overview, victoria, widths = c(2, 1))
```

`patchwork::inset_element()` places the panel over the main map instead. Keep
the rectangle either way, and frame the inset panel in the same colour so the
two read as a pair.

If the report is about people rather than land, consider a tile map or a
population cartogram; a choropleth of rural areas is mostly a map of geography.

## Give each local health area its own colour

Some figures compare areas rather than map a value: which LHA is which on a
reference map, or a line for each LHA over time. `islh_areas()` gives each of
the 14 LHAs a fixed colour, the same in every report, and
`scale_fill_islh_area()` and `scale_colour_islh_area()` apply them.

No palette makes 14 areas identifiable by colour alone, so the colours work in
two levels:

* The health service delivery area picks the colour family: Blue for South,
  Cedar for Central and Thistle for North Vancouver Island. These are the
  brand's low-signal families, meant for organizing content. Red, Green and
  Orange keep their signal meanings.
* The LHA picks the shade within the family. Neighbouring LHAs differ by at
  least 20 values within an HSDA, and South and North never touch, so the
  families that colour-blind readers confuse most never meet on a map.

```{r lha-colours}
islh_areas()
```

The brand standard allows more than two colour families only when colour
carries meaning, and then never as the only cue. So label the areas. On a
chart, one panel per HSDA holds 4 to 6 lines of one family, and a name at the
end of each line does the work a legend would:

```{r lha-lines, fig.height = 7, fig.alt = "Three panels, one per health service delivery area, each showing simulated weekly rates for its local health areas as lines in shades of one colour family, named at the right-hand end."}
weeks <- seq(as.Date("2026-01-04"), by = "week", length.out = 10)
areas <- islh_areas()
rates <- expand.grid(week = weeks, code = areas$code, stringsAsFactors = FALSE)
rates <- merge(rates, areas[c("code", "name", "hsda")])
rates$hsda <- factor(rates$hsda, levels = unique(areas$hsda))
level <- 8 * as.numeric(substr(rates$code, 3, 3))
rates$rate <- 10 + level + 6 * sin(as.numeric(rates$week - min(weeks)) / 20)
last_week <- rates[rates$week == max(weeks), ]

ggplot(rates, aes(week, rate, colour = name)) +
  geom_line(linewidth = 0.9) +
  geom_text(
    data = last_week,
    aes(label = name),
    colour = islh_hex("grey", 15),
    hjust = 0,
    nudge_x = 2,
    size = 3.8
  ) +
  facet_wrap(vars(hsda), ncol = 1) +
  scale_x_date(expand = expansion(mult = c(0.02, 0.45))) +
  scale_colour_islh_area(guide = "none") +
  labs(
    title = "Weekly rate by local health area",
    subtitle = "Simulated values",
    x = NULL,
    y = "Rate per 100,000"
  ) +
  theme_islh()
```

`scale_colour_islh_area()` draws each line in the area's `line_colour`, 10
values darker than its fill, so every line reaches at least 3:1 against the
page, the minimum for a line a reader needs to follow. The fills of the
lightest areas, such as Saanich Peninsula, measure only about 2.3:1 on white,
which is why lines do not use them. Label text sits beside the line in Grey 15,
about 16:1, not in the line's colour: text needs 4.5:1, which not every line
colour reaches.

On a map, `islh_area_labels()` writes each area's code without overlaps.
Eight areas are large enough to hold their code, which sits inside in the
area's `label_colour`: white or Grey 10, whichever contrasts more with the
fill. Every pair reaches at least 4.5:1, so the codes do not rely on being
large. The six
small South and Central Island areas are labelled in a column over the Strait
of Georgia, each joined to its area by a thin line, so no code crosses a
border:

```{r lha-map, fig.height=6, fig.alt="Map of Island Health's 14 local health areas, South Island in shades of blue, Central Island in shades of cedar and North Island in shades of thistle. Eight areas carry their code inside; the six small southern and central areas are labelled in a column to the east, each joined to its area by a thin line."}
lha <- merge(
  islh_example_lha(),
  islh_areas(),
  by.x = "geography_code",
  by.y = "code"
)

ggplot(lha) +
  geom_sf(aes(fill = name), colour = "white", linewidth = 0.3) +
  islh_area_labels() +
  scale_fill_islh_area(name = NULL) +
  coord_islh_map() +
  labs(title = "Island Health local health areas") +
  theme_islh_map(legend = "none")
```

The label positions are fixed for the whole Island. On an inset or a map of
part of it, label the areas yourself.

The scales match the official names or the three-digit codes, so data from
`islh_bc_geography()` works as it comes. A value that matches no LHA, such as
a misspelled name, is drawn grey with a warning naming it.
`scale_fill_islh_area("hsda")` does the same for the three HSDAs. South
Vancouver Island is Blue 45, one step darker than the primary Blue 50: white
text on Blue 50 measures 4.45:1, just short of 4.5:1.

## Colour and contrast

The map palette is a single-hue blue ramp whose steps fall about ten points of
CIE lightness apart. Single hue plus even lightness steps is what keeps a ramp
readable in greyscale and for colour-blind readers, so the brand colours do not
need replacing with Viridis here. Check any palette you build yourself with
`colorspace::specplot()`, view the finished map through `colorspace::deutan()`,
and print it in greyscale once. Internal readers print.

Lightness steps do not replace measuring contrast where it matters. The WCAG
thresholds are ratios, and a ratio just under one fails it:

| What | Minimum against what is behind it |
|---|---|
| Text: labels, codes, legend text | 4.5:1 |
| Large text: 18 pt regular or 14 pt bold, about 24 px or 18.7 px | 3:1 |
| Lines, points and other marks a reader needs to read the figure | 3:1 |

In ggplot2, text `size` is in millimetres: 18 pt is about 6.4 and 14 pt about
4.9. The package's area colours meet these: every `label_colour` on its fill
reaches 4.5:1, and every `line_colour` reaches 3:1 on white. The tests check
the ratios, not the brand values. A map's fills meet each other more than the
page, and are outlined and labelled, so a pale fill beside the white page does
not on its own make the map unreadable.

Use a diverging scale only when the map compares against a baseline, such as a
standardized ratio against the Island Health average. "Where is this worse than
expected" is usually the question, and a ratio answers it more directly than a
rate.

## Alt text

Every map chunk needs `fig-alt` in Quarto. Describe what the map shows, not
that it is a map: the geography, the indicator, and the pattern a reader would
see.

````
```r
#| label: fig-lha-rate
#| fig-alt: >
#|   Choropleth of Island Health local health areas showing influenza
#|   hospitalization rates. Rates are highest in the north Island and
#|   lowest in Greater Victoria.
```
````

## Reproducibility

Capture `attr(population_extract, "bcdata_source")` and the corresponding
attribute on `boundaries` before transforming them. Save a dated extract in
the analysis project when a report must reproduce the exact population or
boundary version used for publication.

The example map in this package keeps its own build record: when the data was
retrieved, the exact addresses it came from, and how it was processed. The
script that builds it also records the hash of each raw response, so a rebuild
after the catalogue is corrected can be told apart from this one:

```{r provenance}
record <- utils::read.csv(
  system.file("extdata", "islh-lha-provenance.csv", package = "islandbrand")
)
record[record$field %in% c("retrieved_utc", "population_year", "crs"), ]
```
