---
title: "Moving from the theme script"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moving from the theme script}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(islandbrand)
```

If your project has an `R/island-health-ggplot-theme.R` in it, it came from a
starter-kit ZIP. This is how to move it to the package.

```{r installed-version}
islh_version()
```

## What changes

The function names you used stay the same. What changes is where they come
from: the package, rather than a copy of the script inside your project. A
fix then reaches your report the next time you update `islandbrand`, instead
of the next time you download a ZIP.

| In the script or starter kit | In the package |
|---|---|
| `theme_islh()`, `islh_setup()`, `islh_flextable()`, the `scale_*_islh()` functions | Kept, with the same names. Check each reference page for new arguments. |
| `source("R/island-health-ggplot-theme.R")` | Replaced by `library(islandbrand)` |
| `islh_theme_version` | Replaced by `islh_version()` |
| `install-packages.R` | Replaced by `islh_install_deps()` |
| A loose `islh-report-reference.docx` and a `reference-doc:` line | Replaced by the `islh-report-docx` format from `islh_use_quarto()` |
| The starter-kit ZIPs | Replaced by `islh_create_report()` |
| Nothing | New: `islh_reset()` and `with_islh()` to undo setup, `islh_check_project()` and `islh_update_project()` to keep project files current |

## Install the package

On an Island Health laptop, install the approved version from its Windows
`.zip`, from your team lead or the
[Releases page](https://github.com/temuulene/islandbrand/releases). R loads
the packages `islandbrand` is built on before any of its functions can run,
so install those first:

```{r install-file, eval=FALSE}
install.packages(
  c("cli", "ggplot2", "rlang", "scales", "systemfonts"),
  type = "binary"
)
install.packages("path/to/islandbrand_x.y.z.zip", repos = NULL)
islandbrand::islh_install_deps("docx") # or "html", or "both"
```

Replace `x.y.z` with the approved version, then restart R. PHASE staff can
instead use `install-phase.R` from the paired `islandepi` release.

On a development machine with access to GitHub:

```{r install-package, eval=FALSE}
install.packages("remotes", type = "binary")
remotes::install_github("temuulene/islandbrand")
```

## Move a project, safely

Work through these in order, and delete nothing until the report renders.

**1. Keep a copy of what you have.** Commit the project to Git, or copy the
folder. The steps below replace files.

**2. Add the format and brand files.** From inside the project:

```{r add-files, eval=FALSE}
islandbrand::islh_use_quarto()
islandbrand::islh_use_brand()
islandbrand::islh_check_project()
```

`islh_check_project()` should report every file up to date. If you had
edited a brand file of your own, keep your edits in the new `_brand.yml`.

**3. Change the setup chunk.** Replace:

```{r old-setup, eval=FALSE}
source("R/island-health-ggplot-theme.R")
islh_setup()
```

with:

```{r new-setup, message=TRUE, warning=FALSE}
library(islandbrand)
islh_setup()
```

In a rendered document, as here, setup detects HTML or Word and sets up the
matching tables. At the console it configures plots only.

**4. Point the front matter at the format.** For a Word report, replace the
`reference-doc:` line with:

```yaml
format: islh-report-docx
```

Use `islh-report-html` for HTML.

**5. Render to every format the report uses**, and compare the output with
the last version you released.

**6. Only then remove what the package replaced:** the copied script, the
`R/` folder if it holds nothing else, and the loose reference document.
Before deleting the loose `islh-report-reference.docx`, check that its
replacement is in `_extensions/islh/islh-report/` and that Git tracks it:

```sh
git ls-files _extensions/islh/islh-report/islh-report-reference.docx
```

That prints the path when the file is tracked. If it prints nothing, see
"Keep the project in Git" in [Writing reports](reports.html).

## Things that are gone

- **`islh_theme_version`** is replaced by `islh_version()`, which reads the
  installed package version. There is no longer a version string inside a
  script to keep in step with anything.
- **`install-packages.R`** is replaced by `islh_install_deps()`. It follows the
  same rules — base R, binary packages, never `pak` — but lives in one place
  rather than in a copy inside every project.
- **The starter-kit ZIPs** are replaced by `islh_create_report()`.

## One thing that may now fail loudly

The package declares minimum versions the script never did:

- **ggplot2 3.5.0** for the colour scales. On an older ggplot2 they used to
  fail inside ggplot2 with `argument "scale_name" is missing`. R now refuses
  to load `islandbrand` with an older ggplot2, and says which version it
  needs.
- **flextable 0.9.10** for `islh_flextable()`. On an older flextable it used to
  fail with `unused argument (repeat_headers = TRUE)`. Now `islh_setup()` and
  every Word table function, including a direct `islh_flextable()` call,
  stop first with a message naming flextable, the version needed and the
  install command.

If a report that used to run now stops with a clear message naming one of
these, that is the version gate doing its job. `islh_install_deps()` will
bring the package up to date; restart R afterwards.

## If you cannot install packages

Some machines will not let you. `islh_check("docx")` or `islh_check("html")`
will tell you what is missing for that format. Your team lead can supply a
built copy of `islandbrand` and the packages it needs, to install from files
as above.
