---
title: "Writing reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
library(islandbrand)
islh_setup(format = "html", quiet = TRUE)
```

## Start a project

```{r create-project, message=TRUE}
project <- file.path(tempdir(), "flu-season-2026")
islandbrand::islh_create_report(project, format = "docx")
```

This article writes to a temporary folder so it can run anywhere; in practice,
give the path where the project should live. The project holds:

```{r project-files}
list.files(project, recursive = TRUE)
```

That is a Quarto project that renders as it stands. Open the `.Rproj`
file, run `islh_install_deps("docx")` once on that machine, then open
`report.qmd` and click **Render**. Replace the example data with your own.

`format` takes `"html"`, `"docx"` or `"both"`. With `"both"`, the report
chooses its table engine at render time, so one file produces either output.

## Add the format to a project you already have

```{r add-format, message=TRUE}
existing <- file.path(tempdir(), "respiratory-summary")
dir.create(existing)

islandbrand::islh_use_quarto(existing)
islandbrand::islh_use_brand(existing)
```

Run from inside your project, both calls need no argument.

The first copies the Island Health Quarto extension into `_extensions/`; the
second copies `_brand.yml` alongside it. Quarto only looks for extensions
inside the project directory, which is why these are copies rather than
references into your R library.

Then name the format:

```yaml
---
title: "Respiratory season summary"
format: islh-report-docx    # or islh-report-html
---
```

## Keeping a project up to date

Because those files are copies, they age. A fix to the Word reference document
reaches your report only when the copy in your project is replaced. Re-copying
blindly is not the answer either, since people edit `_brand.yml` for a
program's own colours and lose that work without warning.

`islh_check_project()` reports where a project stands. It reads only. To show
what it finds, suppose someone added their program's colour to `_brand.yml` and
deleted one of the logos:

```{r edit-project}
cat("# Program accent colour added locally\n",
    file = file.path(project, "_brand.yml"), append = TRUE)
file.remove(file.path(project, "logos", "islh-logo-stacked-white.svg"))
```

```{r check-project, message=TRUE}
islandbrand::islh_check_project(project)
```

Each file is up to date, out of date, missing, or edited here. A project
created by `islh_create_report()` records which version of each file it
installed, in `_islh-manifest.csv`, and that record is what separates a file
that is merely old from one you have changed. Commit it with the rest of the
project.

`islh_update_project()` brings the out-of-date and missing files forward.
Anything you have edited is left alone, and whatever it replaces is copied
into a timestamped folder under `_islh-backup` first.

```{r update-project, message=TRUE}
# See what would change without changing anything.
islandbrand::islh_update_project(project, dry_run = TRUE)

islandbrand::islh_update_project(project)
```

The missing logo is back, and the edited `_brand.yml` is untouched.

If you meant to discard your own edits, `force = TRUE` replaces those files
too. The backup is what makes that recoverable.

A project scaffolded before this feature existed has no manifest, so a file
that differs from the installed version is reported as unverifiable rather than
guessed at, and is left alone until you replace it with `force = TRUE`.

## The setup chunk

Every Island Health report starts the same way:

````
```r
#| label: setup
#| include: false
library(islandbrand)
islh_setup()
```
````

## Figures

Give every figure a caption and alt text. Alt text is what a screen-reader
user gets instead of the picture, so describe what the figure *shows*, not
that it is a bar chart:

````
```{r}
#| label: fig-encounters
#| fig-cap: "Encounters by program"
#| fig-alt: "Primary care had the most encounters at 326, then mental health
#|   at 184, then public health at 79."
#| fig-width: 6
#| fig-height: 3.2
```
````

Reference it in prose with `@fig-encounters` so the numbering stays right when
you move things around.

Here is the complete figure code with package data. This chunk is evaluated
when the website is built, so the figure below is the actual result.

```{r report-figure, fig.alt="Bar chart of example program encounters, with Primary care highest."}
library(ggplot2)

report_data <- islh_example_data()

ggplot(report_data, aes(program, encounters, fill = program)) +
  geom_col(show.legend = FALSE) +
  scale_fill_islh() +
  scale_y_islh_count() +
  labs(
    title = "Encounters by program",
    subtitle = "Illustrative report output",
    x = NULL,
    y = "Encounters"
  )
```

Tables use the same data. Use `islh_gt()` in HTML:

```{r report-table}
islh_gt(
  report_data,
  title = "Program activity",
  source_note = "Source: simulated package data."
)
```

In a dual-format report, select the table engine from the output format:

```{r dual-format-table}
if (knitr::is_html_output()) {
  islh_gt(report_data)
} else {
  islh_flextable(report_data, caption = "Program activity")
}
```

## Grouped tables

When rows fall under headings, such as sites under their health service
delivery area, name the grouping column with `groupname_col`. In HTML,
`rowname_col` also turns a column into row labels, set apart by a thin rule
and keeping its name as a header:

```{r grouped-table}
sites <- data.frame(
  HSDA = c("South", "South", "Central", "Central", "North"),
  Site = c(
    "Rock Bay Landing (Victoria)", "The Harbour (Victoria)",
    "Cowichan (Duncan)", "Port Alberni", "Comox Valley"
  ),
  Opened = c(
    "February 1, 2017", "June 18, 2018", "September 11, 2017",
    "May 1, 2017", "September 1, 2020"
  ),
  Hours = c(
    "7 AM - 10 PM", "8 AM - 7 PM", "8 AM - 8 PM", "8 AM - 8 PM",
    "9:30 AM - 3:30 PM"
  )
)

islh_gt(
  sites,
  groupname_col = "HSDA",
  rowname_col = "Site",
  title = "Site details"
)
```

`islh_flextable()` takes the same `groupname_col`, with the same shaded
heading rows, so a dual-format report groups its table in both outputs. Word
tables have no separate row-label column, so keep that column first. This is
the Word version of the same table:

```{r grouped-word, eval=requireNamespace("flextable", quietly = TRUE) && requireNamespace("officer", quietly = TRUE)}
islh_flextable(sites, groupname_col = "HSDA")
```

In a dual-format report, choose between the two as in the previous section.

## Word output

The Word format uses a reference document built from the official Island
Health letterhead: the logo header, the `Island Health / Page X of Y` footer,
BC Sans, and headings in the letterhead's dark blue.

Two things to know:

- The table of contents renders empty. Word populates it as a field the first
  time the document is opened. That is Word's behaviour, not a fault in the
  template.
- Word locks a file it has open. Close `report.docx` before re-rendering, or
  the render fails on a permission error.

## Troubleshooting

**BC Sans is not being used.** It is not shipped with the package, so it has to
be installed on your machine — once, from
[bcgov/bc-sans](https://github.com/bcgov/bc-sans). `islh_font_family()` reports
what was found. Without it, output falls back to another font; everything else
still works.

**Windows plot pane warns about the font database.** Base graphics devices on
Windows resolve font names through R's own font database rather than through
`systemfonts`. `islh_setup()` registers the family to stop the warnings. If
they persist, set the RStudio graphics backend to AGG under
*Tools > Global Options > General > Graphics*.

**Figures vanish from the rendered document.** On Windows, a path longer than
260 characters silently drops figures. Keep the project somewhere short, not
several folders deep in a synced drive.

**A package is missing or too old.** `islh_check()` names it and gives you the
command. `islh_install_deps()` runs that command for you. Both use base R's
`install.packages()` with `type = "binary"`, because Island Health laptops
block installers that unpack with their own helper binary and have no compiler
for a source build.

## Before anything leaves your team

Reports built this way are ordinary files with no access controls. Never put
identifiable information in one, and suppress small cells before sharing:

```r
counts$cases <- islandepi::islh_suppress(counts$cases, threshold = 5)
```

`islandepi::islh_suppress_table()` does a whole table, and its `complementary`
argument
handles the case where a visible total would let a reader recover a suppressed
cell by subtraction. The threshold is yours to set — check the rule your
release is governed by.
