---
title: "Surveillance figures and epidemic curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surveillance figures and epidemic curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  fig.retina = 2
)
library(islandbrand)
islh_setup(format = "html", quiet = TRUE)
```

`islh_epi_curve()` creates a complete Island Health surveillance figure from
prepared counts. It returns an ordinary ggplot, so titles, annotations, scales
and additional layers can be changed with ggplot2.

Counting and plotting are deliberately separate. When starting from a line
list, create complete counts with `islandepi::islh_count_events()` and pass its
result to this function.

## Routine weekly surveillance

The input needs a date column and a whole-count column. This example is already
aggregated by week and source.

```{r weekly-data}
weeks <- seq(as.Date("2026-01-04"), by = "week", length.out = 14)

weekly_counts <- data.frame(
  period_start = rep(weeks, each = 2),
  source = rep(c("Community", "Facility"), times = length(weeks)),
  count = c(
    1, 0, 2, 1, 2, 1, 4, 2, 6, 3, 9, 4, 8, 3,
    6, 2, 5, 2, 4, 1, 3, 1, 2, 1, 1, 0, 1, 0
  )
)

head(weekly_counts)
```

```{r weekly-curve, fig.alt="Stacked weekly epidemic curve rising to a peak and declining, split between community and facility reports."}
islh_epi_curve(
  weekly_counts,
  date = period_start,
  count = count,
  fill = source,
  labels = "total",
  title = "Reported cases by week",
  subtitle = "Illustrative surveillance data",
  caption = "Source: simulated package example."
)
```

The bar width is inferred from the distance between dates. Weekly observations
therefore fill most of the week instead of rendering as one-day-wide bars.

Use `position = "dodge"` when the group comparison matters more than the total:

```{r dodged-curve, fig.alt="Grouped weekly bars comparing community and facility reports."}
islh_epi_curve(
  weekly_counts,
  date = period_start,
  count = count,
  fill = source,
  position = "dodge",
  title = "Weekly reports by source"
)
```

## One rectangle per case

For a small outbreak, `style = "cases"` outlines each case. The default remains
ordinary bars because routine surveillance counts can be too large to expand
into thousands of rectangles.

```{r case-curve, fig.alt="Small epidemic curve with one outlined rectangle for every case."}
small_outbreak <- data.frame(
  onset_date = rep(as.Date("2026-03-01") + 0:9, each = 2),
  exposure = rep(c("Known exposure", "Under investigation"), 10),
  count = c(0, 1, 1, 1, 2, 1, 4, 2, 6, 3, 5, 2, 3, 2, 2, 1, 1, 1, 0, 1)
)

islh_epi_curve(
  small_outbreak,
  date = onset_date,
  count = count,
  fill = exposure,
  style = "cases",
  title = "Cases by symptom-onset date",
  y = "Cases"
)
```

The function stops before creating more than 50,000 rectangles. Increase
`max_cases` only when an individual-case display is genuinely required.

## Historical reference range

Supply a separate data frame with dates and lower and upper limits. A
`reference_mean` column adds a dashed centre line.

```{r reference-curve, fig.alt="Weekly epidemic curve over a grey historical reference range and dashed historical mean."}
totals <- aggregate(count ~ period_start, weekly_counts, sum)

historical_reference <- data.frame(
  period_start = weeks,
  lower_limit = pmax(0, c(0, 0, 1, 1, 2, 2, 3, 2, 2, 1, 1, 0, 0, 0)),
  upper_limit = c(4, 4, 5, 6, 7, 9, 10, 9, 8, 7, 6, 5, 4, 4),
  reference_mean = c(2, 2, 3, 3, 4, 5, 6, 5, 5, 4, 3, 3, 2, 2)
)

islh_epi_curve(
  totals,
  date = period_start,
  count = count,
  reference = historical_reference,
  title = "Weekly cases compared with historical activity",
  subtitle = "The grey band is descriptive, not an outbreak declaration"
)
```

The function draws the reference but does not interpret it. Add an annotation
or signal layer only after applying the surveillance program's approved rule.

## Facet by geography

```{r faceted-curve, fig.alt="Two small-multiple epidemic curves for North and South regions."}
regional <- rbind(
  transform(weekly_counts, region = "North", count = pmax(0, count - 1)),
  transform(weekly_counts, region = "South")
)

islh_epi_curve(
  regional,
  date = period_start,
  count = count,
  fill = source,
  facet = region,
  title = "Reported cases by region"
)
```

For many geographic groups, faceting is generally easier to read than assigning
every geography a different colour.

## Start from event-level data

The complete two-package workflow is:

```{r two-package-workflow, eval=FALSE}
counts <- islandepi::islh_count_events(
  linelist,
  date = surveillance_date,
  id = investigation_id,
  by = c(condition, region),
  interval = "epiweek",
  fill = TRUE,
  include_partial = FALSE
)

islh_epi_curve(
  counts,
  date = period_start,
  count = count,
  fill = condition,
  facet = region,
  title = "Weekly communicable disease reports"
)
```

The plotting package does not depend on `islandepi`; prepared counts from
another validated workflow work equally well.
