---
title: "Surveillance figures and epidemic curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surveillance figures and epidemic curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  fig.retina = 2
)
library(islandbrand)
islh_setup(format = "html", quiet = TRUE)
```

`islh_epi_curve()` creates a complete Island Health surveillance figure from
prepared counts. It returns an ordinary ggplot, so titles, annotations, scales
and additional layers can be changed with ggplot2.

Counting and plotting are deliberately separate. When starting from a line
list, create complete counts with `islandepi::islh_count_events()` and pass its
result to this function.

## Routine weekly surveillance

The input needs a date column and a whole-count column. This example is already
aggregated by week and source.

```{r weekly-data}
weeks <- seq(as.Date("2026-01-04"), by = "week", length.out = 14)

weekly_counts <- data.frame(
  period_start = rep(weeks, each = 2),
  source = rep(c("Community", "Facility"), times = length(weeks)),
  count = c(
    1, 0, 2, 1, 2, 1, 4, 2, 6, 3, 9, 4, 8, 3,
    6, 2, 5, 2, 4, 1, 3, 1, 2, 1, 1, 0, 1, 0
  )
)

head(weekly_counts)
```

```{r weekly-curve, fig.alt="Stacked weekly epidemic curve rising to a peak and declining, split between community and facility reports."}
islh_epi_curve(
  weekly_counts,
  date = period_start,
  count = count,
  fill = source,
  labels = "total",
  title = "Reported cases by week",
  subtitle = "Illustrative surveillance data",
  caption = "Source: simulated package example."
)
```

The bar width is inferred from the distance between dates. Weekly observations
therefore fill most of the week instead of rendering as one-day-wide bars.

Use `position = "dodge"` when the group comparison matters more than the total:

```{r dodged-curve, fig.alt="Grouped weekly bars comparing community and facility reports."}
islh_epi_curve(
  weekly_counts,
  date = period_start,
  count = count,
  fill = source,
  position = "dodge",
  title = "Weekly reports by source"
)
```

## One rectangle per case

For a small outbreak, `style = "cases"` outlines each case. The default remains
ordinary bars because routine surveillance counts can be too large to expand
into thousands of rectangles.

```{r case-curve, fig.alt="Small epidemic curve with one outlined rectangle for every case."}
small_outbreak <- data.frame(
  onset_date = rep(as.Date("2026-03-01") + 0:9, each = 2),
  exposure = rep(c("Known exposure", "Under investigation"), 10),
  count = c(0, 1, 1, 1, 2, 1, 4, 2, 6, 3, 5, 2, 3, 2, 2, 1, 1, 1, 0, 1)
)

islh_epi_curve(
  small_outbreak,
  date = onset_date,
  count = count,
  fill = exposure,
  style = "cases",
  title = "Cases by symptom-onset date",
  y = "Cases"
)
```

The function stops before creating more than 50,000 rectangles. Increase
`max_cases` only when an individual-case display is genuinely required.

## Historical reference range

Supply a separate data frame with dates and lower and upper limits. A
`reference_mean` column adds a dashed centre line.

```{r reference-curve, fig.alt="Weekly epidemic curve over a grey historical reference range and dashed historical mean."}
totals <- aggregate(count ~ period_start, weekly_counts, sum)

historical_reference <- data.frame(
  period_start = weeks,
  lower_limit = pmax(0, c(0, 0, 1, 1, 2, 2, 3, 2, 2, 1, 1, 0, 0, 0)),
  upper_limit = c(4, 4, 5, 6, 7, 9, 10, 9, 8, 7, 6, 5, 4, 4),
  reference_mean = c(2, 2, 3, 3, 4, 5, 6, 5, 5, 4, 3, 3, 2, 2)
)

islh_epi_curve(
  totals,
  date = period_start,
  count = count,
  reference = historical_reference,
  title = "Weekly cases compared with historical activity",
  subtitle = "The grey band is descriptive, not an outbreak declaration"
)
```

The function draws the reference but does not interpret it. Add an annotation
or signal layer only after applying the surveillance program's approved rule.

## Showing uncertainty

A rate or percentage in a figure needs its confidence interval as much as one
in a table. Without it, a reader cannot tell a rate based on 3 cases from one
based on 300. The islandepi article
[Why every estimate needs an interval](https://temuulene.github.io/islandepi/articles/intervals.html)
works through why with examples. This section shows how to draw intervals in
the Island Health style.

Calculate the rates and their intervals first.
`islandepi::islh_crude_rate()`, `islandepi::islh_proportion()` and
`islandepi::islh_dsr()` all return `lower` and `upper` columns beside the
estimate:

```{r interval-workflow, eval=FALSE}
area_rates <- cbind(
  areas["lha"],
  islandepi::islh_crude_rate(areas$cases, areas$population)
)
```

### Areas or groups

Draw each area as a dot with a line through it for the interval. Sort areas by
name or geography, not by rate: sorting by rate puts the smallest, least certain
areas at both ends. A dashed line for the overall rate gives readers a
reference point.

```{r area-intervals, fig.height=4.5, fig.alt="Dot-and-whisker chart of annual rates per 100,000 for eight local health areas in alphabetical order, with a dashed line at the Island Health rate of 49.5. Vancouver Island West has the highest rate, 134, but its interval runs from 28 to 392. Cowichan Valley West has the lowest, 14, with an interval from 0.3 to 76. Every interval crosses the dashed line."}
area_rates <- data.frame(
  lha = c(
    "Alberni/Clayoquot", "Comox Valley", "Cowichan Valley North",
    "Cowichan Valley South", "Cowichan Valley West", "Greater Campbell River",
    "Oceanside", "Vancouver Island West"
  ),
  rate = c(30.3, 44.9, 76.0, 60.5, 13.6, 52.6, 45.6, 134.3),
  lower = c(15.1, 31.5, 44.2, 43.2, 0.3, 34.4, 29.5, 27.7),
  upper = c(54.2, 62.2, 121.7, 82.4, 76.0, 77.1, 67.3, 392.4)
)
area_rates$lha <- factor(area_rates$lha, levels = rev(area_rates$lha))

ggplot2::ggplot(area_rates, ggplot2::aes(x = rate, y = lha)) +
  ggplot2::geom_vline(
    xintercept = 49.5,
    linetype = "dashed",
    colour = islh_hex("grey", 50)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(xmin = lower, xmax = upper),
    colour = islh_brand("primary")
  ) +
  ggplot2::labs(
    x = "Cases per 100,000",
    y = NULL,
    title = "Annual rate by local health area",
    subtitle = "Dashed line: Island Health overall",
    caption = "Lines show 95% confidence intervals. Illustrative data."
  )
```

Say what the lines show, in the caption or the axis title. Avoid bar charts of
rates without intervals: every bar looks equally certain. If a bar chart is
needed, add `ggplot2::geom_errorbar()` with `width = 0.2`.

### Trends over time

Draw the interval as a shaded band behind the line. The band widens where
counts are low, which is exactly where month-to-month changes mean least.

```{r trend-intervals, fig.alt="Line chart of a monthly rate per 100,000 from April 2025 to March 2026 with a pale blue 95% confidence band. The rate falls from about 11 in April to about 4 in September, then rises to a peak of about 22 in February. The band is wide throughout. The summer bands overlap each other, while the February band sits entirely above the summer estimates."}
monthly <- data.frame(
  month = seq(as.Date("2025-04-01"), by = "month", length.out = 12),
  rate = c(11.2, 8.7, 7.5, 6.2, 5.0, 3.7, 5.0, 7.5, 11.2, 16.2, 22.5, 18.7),
  lower = c(5.1, 3.5, 2.7, 2.0, 1.4, 0.8, 1.4, 2.7, 5.1, 8.6, 13.3, 10.5),
  upper = c(21.3, 18.0, 16.3, 14.6, 12.8, 10.9, 12.8, 16.3, 21.3, 27.8, 35.5, 30.9)
)

ggplot2::ggplot(monthly, ggplot2::aes(x = month, y = rate)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = lower, ymax = upper),
    fill = islh_hex("blue", 85)
  ) +
  ggplot2::geom_line(colour = islh_brand("primary"), linewidth = 1) +
  ggplot2::geom_point(colour = islh_brand("primary")) +
  ggplot2::scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
  ggplot2::expand_limits(y = 0) +
  ggplot2::labs(
    x = NULL,
    y = "Cases per 100,000",
    title = "Monthly rate in one local health area",
    caption = "Shaded band: 95% confidence interval. Illustrative data."
  )
```

The summer rates, 4 to 8 per 100,000, each rest on 3 to 6 cases, so their
bands are wide and overlap one another: the dip from June to September is not
clear evidence of a fall. The winter rise is clearer. By February the whole
band sits above every summer estimate.

### Tables

Put the interval beside the estimate, as `30.3 (15.1 to 54.2)`, and name the
method and confidence level in the table notes. Use "to" rather than a hyphen,
which reads like a minus sign. Format the columns, then pass the data frame to
`islh_gt()` or `islh_flextable()`.

## Facet by geography

```{r faceted-curve, fig.alt="Two small-multiple epidemic curves for North and South regions."}
regional <- rbind(
  transform(weekly_counts, region = "North", count = pmax(0, count - 1)),
  transform(weekly_counts, region = "South")
)

islh_epi_curve(
  regional,
  date = period_start,
  count = count,
  fill = source,
  facet = region,
  title = "Reported cases by region"
)
```

For many geographic groups, faceting is generally easier to read than assigning
every geography a different colour.

## Start from event-level data

The complete two-package workflow is:

```{r two-package-workflow, eval=FALSE}
counts <- islandepi::islh_count_events(
  linelist,
  date = surveillance_date,
  id = investigation_id,
  by = c(condition, region),
  interval = "epiweek",
  fill = TRUE,
  include_partial = FALSE
)

islh_epi_curve(
  counts,
  date = period_start,
  count = count,
  fill = condition,
  facet = region,
  title = "Weekly communicable disease reports"
)
```

The plotting package does not depend on `islandepi`; prepared counts from
another validated workflow work equally well.
