---
title: "Surveillance figures and epidemic curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surveillance figures and epidemic curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  fig.retina = 2
)
library(islandbrand)
islh_setup(format = "html", quiet = TRUE)
```

`islh_epi_curve()` creates a complete Island Health surveillance figure from
prepared counts. It returns an ordinary ggplot, so titles, annotations, scales
and additional layers can be changed with ggplot2.

Counting and plotting are deliberately separate. When starting from a line
list, create complete counts with `islandepi::islh_count_events()` and pass its
result to this function.

## Routine weekly surveillance

The input needs a date column and a whole-count column. This example is already
aggregated by week and source.

```{r weekly-data}
weeks <- seq(as.Date("2026-01-04"), by = "week", length.out = 14)

weekly_counts <- data.frame(
  period_start = rep(weeks, each = 2),
  source = rep(c("Community", "Facility"), times = length(weeks)),
  count = c(
    1, 0, 2, 1, 2, 1, 4, 2, 6, 3, 9, 4, 8, 3,
    6, 2, 5, 2, 4, 1, 3, 1, 2, 1, 1, 0, 1, 0
  )
)

head(weekly_counts)
```

```{r weekly-curve, fig.alt="Stacked weekly epidemic curve rising to a peak and declining, split between community and facility reports."}
islh_epi_curve(
  weekly_counts,
  date = period_start,
  count = count,
  fill = source,
  labels = "total",
  title = "Reported cases by week",
  subtitle = "Illustrative surveillance data",
  caption = "Source: simulated package example."
)
```

The bar width is inferred from the distance between dates: nine tenths of the
shortest gap, so weekly bars fill most of the week. Start dates alone cannot
say more than that:

- With a single period there is no gap to measure. Counts from
  `islandepi::islh_count_events()` record their interval, which sets the
  width; other data gets 0.9 days.
- Monthly bars are sized to the shortest month, 28 days, and a missing period
  looks like a gap the data never had.

For irregular periods, set `bar_width` in days yourself.

Use `position = "dodge"` when the group comparison matters more than the total:

```{r dodged-curve, fig.alt="Grouped weekly bars comparing community and facility reports."}
islh_epi_curve(
  weekly_counts,
  date = period_start,
  count = count,
  fill = source,
  position = "dodge",
  title = "Weekly reports by source"
)
```

## One rectangle per case

For a small outbreak, `style = "cases"` outlines each case. The default remains
ordinary bars because routine surveillance counts can be too large to expand
into thousands of rectangles.

```{r case-curve, fig.alt="Small epidemic curve with one outlined rectangle for every case."}
small_outbreak <- data.frame(
  onset_date = rep(as.Date("2026-03-01") + 0:9, each = 2),
  exposure = rep(c("Known exposure", "Under investigation"), 10),
  count = c(0, 1, 1, 1, 2, 1, 4, 2, 6, 3, 5, 2, 3, 2, 2, 1, 1, 1, 0, 1)
)

islh_epi_curve(
  small_outbreak,
  date = onset_date,
  count = count,
  fill = exposure,
  style = "cases",
  title = "Cases by symptom-onset date",
  y = "Cases"
)
```

The function stops before creating more than 50,000 rectangles. Increase
`max_cases` only when an individual-case display is genuinely required.

## Historical reference range

Supply a separate data frame with dates and lower and upper limits. A
`reference_mean` column adds a dashed centre line.

```{r reference-curve, fig.alt="Weekly epidemic curve over a grey historical reference range and dashed historical mean."}
totals <- aggregate(count ~ period_start, weekly_counts, sum)

historical_reference <- data.frame(
  period_start = weeks,
  lower_limit = pmax(0, c(0, 0, 1, 1, 2, 2, 3, 2, 2, 1, 1, 0, 0, 0)),
  upper_limit = c(4, 4, 5, 6, 7, 9, 10, 9, 8, 7, 6, 5, 4, 4),
  reference_mean = c(2, 2, 3, 3, 4, 5, 6, 5, 5, 4, 3, 3, 2, 2)
)

islh_epi_curve(
  totals,
  date = period_start,
  count = count,
  reference = historical_reference,
  title = "Weekly cases compared with historical activity",
  subtitle = "The grey band is descriptive, not an outbreak declaration"
)
```

The function draws the reference but does not interpret it. Add an annotation
or signal layer only after applying the surveillance program's approved rule.

The reference needs one row per date. A reference with two rows for a date,
such as one per source, would be drawn as one line zigzagging between them,
so it stops instead:

```{r reference-duplicate, error=TRUE}
by_source <- merge(
  historical_reference,
  data.frame(source = c("Community", "Facility"))
)

islh_epi_curve(
  totals,
  date = period_start,
  count = count,
  reference = by_source
)
```

Filter the reference to the series the bars show. In a faceted figure, a
reference with the facet column gives each panel its own band, one row per
date and panel; a reference without it is drawn in every panel.

## Showing uncertainty

A rate or percentage in a figure needs its confidence interval as much as one
in a table. Without it, a reader cannot tell a rate based on 3 cases from one
based on 300. The islandepi article
[Why every estimate needs an interval](https://temuulene.github.io/islandepi/articles/intervals.html)
works through why with examples. This section shows how to draw intervals in
the Island Health style.

Calculate the rates and their intervals first.
`islandepi::islh_crude_rate()`, `islandepi::islh_proportion()` and
`islandepi::islh_dsr()` all return `lower` and `upper` columns beside the
estimate:

```{r interval-workflow, eval=FALSE}
area_rates <- cbind(
  areas["lha"],
  islandepi::islh_crude_rate(areas$cases, areas$population)
)
```

### Areas or groups

Draw each area as a dot with a line through it for the interval. Sort areas by
name or geography, not by rate: sorting by rate puts the smallest, least certain
areas at both ends. A dashed line for the overall rate gives readers a
reference point.

```{r area-intervals, fig.height=4.5, fig.alt="Dot-and-whisker chart of annual rates per 100,000 for eight local health areas in alphabetical order, with a dashed line at the Island Health rate of 49.5. Vancouver Island West has the highest rate, 134, but its interval runs from 28 to 392. Cowichan Valley West has the lowest, 14, with an interval from 0.3 to 76. Every interval crosses the dashed line."}
area_rates <- data.frame(
  lha = c(
    "Alberni/Clayoquot", "Comox Valley", "Cowichan Valley North",
    "Cowichan Valley South", "Cowichan Valley West", "Greater Campbell River",
    "Oceanside", "Vancouver Island West"
  ),
  rate = c(30.3, 44.9, 76.0, 60.5, 13.6, 52.6, 45.6, 134.3),
  lower = c(15.1, 31.5, 44.2, 43.2, 0.3, 34.4, 29.5, 27.7),
  upper = c(54.2, 62.2, 121.7, 82.4, 76.0, 77.1, 67.3, 392.4)
)
area_rates$lha <- factor(area_rates$lha, levels = rev(area_rates$lha))

ggplot2::ggplot(area_rates, ggplot2::aes(x = rate, y = lha)) +
  ggplot2::geom_vline(
    xintercept = 49.5,
    linetype = "dashed",
    colour = islh_hex("grey", 50)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(xmin = lower, xmax = upper),
    colour = islh_brand("primary")
  ) +
  ggplot2::labs(
    x = "Cases per 100,000",
    y = NULL,
    title = "Annual rate by local health area",
    subtitle = "Dashed line: Island Health overall",
    caption = "Lines show 95% confidence intervals. Illustrative data."
  )
```

Say what the lines show, in the caption or the axis title. Avoid bar charts of
rates without intervals: every bar looks equally certain. If a bar chart is
needed, add `ggplot2::geom_errorbar()` with `width = 0.2`.

### Trends over time

Draw the interval as a shaded band behind the line. The band widens where
counts are low, which is exactly where month-to-month changes mean least.

```{r trend-intervals, fig.alt="Line chart of a monthly rate per 100,000 from April 2025 to March 2026 with a pale blue 95% confidence band. The rate falls from about 11 in April to about 4 in September, then rises to a peak of about 22 in February. The band is wide throughout. The summer bands overlap each other, while the February band sits entirely above the summer estimates."}
monthly <- data.frame(
  month = seq(as.Date("2025-04-01"), by = "month", length.out = 12),
  rate = c(11.2, 8.7, 7.5, 6.2, 5.0, 3.7, 5.0, 7.5, 11.2, 16.2, 22.5, 18.7),
  lower = c(5.1, 3.5, 2.7, 2.0, 1.4, 0.8, 1.4, 2.7, 5.1, 8.6, 13.3, 10.5),
  upper = c(21.3, 18.0, 16.3, 14.6, 12.8, 10.9, 12.8, 16.3, 21.3, 27.8, 35.5, 30.9)
)

ggplot2::ggplot(monthly, ggplot2::aes(x = month, y = rate)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = lower, ymax = upper),
    fill = islh_hex("blue", 85)
  ) +
  ggplot2::geom_line(colour = islh_brand("primary"), linewidth = 1) +
  ggplot2::geom_point(colour = islh_brand("primary")) +
  ggplot2::scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y") +
  ggplot2::expand_limits(y = 0) +
  ggplot2::labs(
    x = NULL,
    y = "Cases per 100,000",
    title = "Monthly rate in one local health area",
    caption = "Shaded band: 95% confidence interval. Illustrative data."
  )
```

The summer rates, 4 to 8 per 100,000, each rest on 3 to 6 cases, so their
bands are wide: these estimates cannot say whether the dip from June to
September is real. The winter rise is larger than that uncertainty. By
February the whole band sits above every summer estimate.

Read the bands as a guide, not a test. Two intervals that overlap do not show
that two rates are the same, and two rates can differ clearly even when their
intervals overlap. To compare two periods or two areas, calculate the rate
ratio or difference with its own interval, as the islandepi article shows.
Months next to each other share a season and often the same outbreak, so they
are not independent either; a trend question needs a model that allows for
that.

### Tables

Put the interval beside the estimate, as `30.3 (15.1 to 54.2)`, and name the
method and confidence level in the table notes, such as "95% confidence
intervals, exact Poisson method". Keep the three together wherever the
estimate goes, so a figure caption says the same as the table beside it. Use
"to" rather than a hyphen, which reads like a minus sign. Format the columns,
then pass the data frame to `islh_gt()` or `islh_flextable()`.

## Facet by geography

```{r faceted-curve, fig.alt="Two small-multiple epidemic curves for North and South regions."}
regional <- rbind(
  transform(weekly_counts, region = "North", count = pmax(0, count - 1)),
  transform(weekly_counts, region = "South")
)

islh_epi_curve(
  regional,
  date = period_start,
  count = count,
  fill = source,
  facet = region,
  title = "Reported cases by region"
)
```

For many geographic groups, faceting is generally easier to read than assigning
every geography a different colour.

### Quiet regions and unknown regions

A region with no cases is a result, not a gap. Keep its rows with a count of
zero and it keeps its panel, in bars and in case tiles alike:

```{r quiet-region, fig.alt="Two small-multiple case-tile epidemic curves. The Active panel shows cases in three of four days. The Quiet panel has the same dates and no tiles."}
outbreak_by_region <- data.frame(
  onset_date = rep(as.Date("2026-03-01") + 0:3, times = 2),
  region = rep(c("Active", "Quiet"), each = 4),
  count = c(0, 2, 3, 1, 0, 0, 0, 0)
)

islh_epi_curve(
  outbreak_by_region,
  date = onset_date,
  count = count,
  facet = region,
  style = "cases",
  date_breaks = "1 day",
  title = "Cases by symptom-onset date and region"
)
```

A row with no region is drawn in a panel of its own, labelled `NA`, and its
totals are labelled like any other panel's:

```{r unknown-region, fig.alt="Three small-multiple weekly epidemic curves with total labels: North, South and NA, which holds the cases with no region recorded."}
with_unknown <- data.frame(
  period_start = rep(weeks[1:4], times = 3),
  region = rep(c("North", "South", NA), each = 4),
  count = c(3, 5, 4, 2, 6, 8, 7, 5, 1, 0, 2, 1)
)

islh_epi_curve(
  with_unknown,
  date = period_start,
  count = count,
  facet = region,
  labels = "total",
  date_breaks = "2 weeks",
  title = "Reported cases by week and region"
)
```

Before publishing, give that panel a name readers understand:

```{r unknown-region-name, eval=FALSE}
with_unknown$region[is.na(with_unknown$region)] <- "Region not recorded"
```

A period with no row at all is different from a zero: the function cannot
tell an absent row from a week nobody reported on, so it leaves the space
empty. Fill in zeros only where reporting is known to be complete, as the next
section explains.

## Start from event-level data

The complete two-package workflow is:

```{r two-package-workflow, eval=FALSE}
counts <- islandepi::islh_count_events(
  linelist,
  date = surveillance_date,
  id = investigation_id,
  by = c(condition, region),
  interval = "epiweek",
  # The weeks reporting is confirmed complete for, not the first and last
  # event: the first case is not when surveillance started.
  from = "2025-08-31",
  to = "2026-02-28",
  # Every condition and region expected, including any with no events.
  groups = expected_groups,
  fill = TRUE,
  include_partial = FALSE
)

islh_epi_curve(
  counts,
  date = period_start,
  count = count,
  fill = condition,
  facet = region,
  title = "Weekly communicable disease reports"
)
```

`fill = TRUE` turns a week with no events into a count of zero. That is right
only inside `from` and `to`, and only for groups and weeks where every source
was reporting. A week when a feed was down, or a site had not yet joined, is
unknown rather than zero: leave it out of the window, or show it separately.
The islandepi surveillance guide covers coverage checks before counting.

The plotting package does not depend on `islandepi`; prepared counts from
another validated workflow work equally well.
